\setchapterstyle{kao}
\setchapterpreamble[u]{\margintoc}

\chapter{MIDI}
\labch{midi}

\section{What is MIDI}

\glsreset{midi}\gls{midi} is a protocol for exchanging musical events between devices like synthesizers, keyboards or other electrical instruments. It first appeared as \gls{midi} 1.0 in 1982 and is slowly being replaced by \gls{midi} 2.0, which was introduced in 2020.

Instead of describing the sound of an audio, like, e.g., MP3 does, MIDI describes every tone's pitch, volume, and duration. This enables high flexibility for music production while still being easy to use. The Interrupter will act as a very simple synthesizer that controls the tesla coil to create the described tones for this project. 

\section{Standard MIDI File}

All the information in the following section was taken from \sidecite{midi-spec}.

Unlike MIDI streams, which connects two or more devices and lets them communicate in real time, the MIDI-Interrupter will use so-called \glspl{smf}, which are binary files holding \gls{midi} data.

\glspl{smf} are build up of a header chunk and one or more track chunks. The header chunk stores the file format, number of track chunks, and the division\sidenotes{The time resolution of the file}. The track chunks then hold the actual \gls{midi} data. How these track chunks have to be read depends on the file format.

\begin{tabular}{cll@{}}
    \midrule
    0 & Single Track File Format   & \\
    1 & Multiple Track File Format & All tracks are played simultaneously\\
    2 & Multiple Song File Format  & All tracks are played after one another\\
    \midrule
\end{tabular}

The most common format is the Multiple Track File Format. For songs containing multiple instruments, each instrument usually has its own track. In the same way, piano pieces typically consist of two tracks - one for the left hand and one for the right hand.

The track itself is a series of delta times with a following \gls{midi}, Meta, or Sysex event.

\subsection{Events}

Events are instructions for reconstructing the sound of a song. They always consist of an identifier and some amount of data. This amount depends on the event, but some events also have a variable amount of data, where the length is given as the first argument. The three types of events are \gls{midi} Events, Meta Events, and Sysex Events

\subsubsection{MIDI Events}

\gls{midi}-events contain the actual music information, like \cinl{NOTE_ON} or \cinl{NOTE_OFF}. They consist of a 4-bit identifier, a 4-bit channel, and (in most cases) 2 bytes of data. For the example of \cinl{NOTE_ON} and \cinl{NOTE_OFF}, the first byte of the data contains the note number, and the second byte contains the velocity, which can be interpreted as its loudness.

\begin{figure}[h!]
  \centering
  \begin{bytefield}[bitwidth=1em]{26}
    \bitbox{1}[bgcolor=gr10]{1} &
    \bitbox{1}[bgcolor=white]{0} &
    \bitbox{1}[bgcolor=white]{0} &
    \bitbox{1}[bgcolor=white]{1} &
    \bitbox{4}{channel} &
    \bitbox[]{1}{} &
    \bitbox{1}[bgcolor=gr10]{0} &
    \bitbox{7}{note number} &
    \bitbox[]{1}{} &
    \bitbox{1}[bgcolor=gr10]{0} &
    \bitbox{7}{velocity}\\
  \end{bytefield}
  \caption{Structure of a NOTE\_ON event}
  \label{fig:note-on}
\end{figure}

The 7-bit note number can range from 0 to 127, corresponding to the musical notes C-1 to G9. For reference, a typical 88-key piano ranges from A0 to C8 (note numbers 21 to 108). To convert this note number into a usable frequency, equation \ref{eq:note-number} can be used.

\begin{equation}\label{eq:note-number}
    f = 440 \cdot 2^{\frac{n - 69}{12}}
\end{equation}

\subsubsection{Meta Events}

Non-musical information like song names, lyric texts, tempo settings, or end of track are contained in Meta events. The most common Meta events are \cinl{SET_TEMPO} and \cinl{END_OF_TRACK}, which have to be every track's last event. Every Meta event consists of a \cinl{0xFF}, an identifier, and a variable amount of data.

\subsection{Delta Time}

Delta times specify the time offset to the previous event. They are encoded as Variable Length Quantities, which means they can take up one to four bytes, depending on their size. The MSB % acronym
of each byte shows if another byte follows, and the remaining 7 bits are the delta time. An example implementation for decoding a Variable Length Quantity in C++ can be seen in the listing \ref{lst:delta-time}.

\begin{lstlisting}[caption=Decoding the Delta Time, label=lst:delta-time]
uint32_t deltaTime = 0;
for (uint8_t i = 0; i < 4; i++) {
    uint8_t byte = nextByte();
    deltaTime = (deltaTime << 7) | (byte & 0x7F);
    if(~byte & 0x80) {
         break;
    }
}
\end{lstlisting}

The delta time is the number of ticks, which by itself is not very useful. To convert the delta time into seconds, both the division and the tempo are needed. The division is part of the header chunk, and the tempo can be set with a Meta event. Equation \ref{eq:delta-time} shows how those values can be used to calculate the actual time.

\begin{equation}\label{eq:delta-time}
    \textnormal{time [\textmu s]} = \frac{{\textnormal{tempo} \left[ \frac{\textnormal{\textmu s}}{\textnormal{quarter note}} \right] \cdot \textnormal{delta time } \left[\frac{\textnormal{ticks}}{1}\right]}}{\textnormal{division}  \left[ \frac{\textnormal{ticks}}{\textnormal{quarter note}} \right]}
\end{equation}

\subsection{Running Status}

One thing all event types have in common is that their first identifier byte has a \gls{msb} of 1, and their first argument byte has an \gls{msb} of 0. This way, in case the identifier is omitted, the argument is still clearly recognizable as such and cannot be mistaken for the identifier. The running status scheme now allows it to omit the identifier of an event if it is of the same type as the previous event. For example, after a \newline NOTE\_ON event, consisting of a delta time, an identifier, and its argument, a following NOTE\_ON event could only consist of a delta time followed by the argument.

Only by combining this with another trick, which is to turn all NOTE\_OFF events into NOTE\_ON events with a velocity of zero, does this become useful. Now, for a simple \gls{midi} file, almost the whole file would consist of NOTE\_ON events stringed together, and only the first one would need an identifier.

\begin{figure}[h!]
    \centering
    \begin{minipage}{0.45\textwidth}
  \begin{bytefield}[bitwidth=1.5mm]{32}
  \bitbox{8}{\small \(\Delta\)Time}&
  \bitbox{8}{\small N\_ON}&
  \bitbox{8}{\small Note}&
  \bitbox{8}{\small Velocity}\\
  \bitbox{8}{\small \(\Delta\)Time}&
  \bitbox{8}{\small N\_OFF}&
  \bitbox{8}{\small Note}&
  \bitbox{8}{\small Velocity}\\
  \bitbox{8}{\small \(\Delta\)Time}&
  \bitbox{8}{\small N\_ON}&
  \bitbox{8}{\small Note}&
  \bitbox{8}{\small Velocity}\\
  \bitbox{8}{\small \(\Delta\)Time}&
  \bitbox{8}{\small N\_OFF}&
  \bitbox{8}{\small Note}&
  \bitbox{8}{\small Velocity}\\
  \bitbox[]{32}{\(\vdots\)}
\end{bytefield}
\end{minipage}%
    \begin{minipage}{0.1\textwidth}
      \begin{center}
        \raisebox{5mm}{\resizebox{5mm}{!}{\(\Rightarrow\)}}
      \end{center}
\end{minipage}%
    \begin{minipage}{0.45\textwidth}
    \begin{bytefield}[bitwidth=1.5mm]{32}
      \bitbox{8}{\small \(\Delta\)Time}&
      \bitbox{8}{\small N\_ON}&
      \bitbox{8}{\small Note}&
      \bitbox{8}{\small Velocity}\\
      \bitbox{8}{\small \(\Delta\)Time}&
      \bitbox{8}{\small Note}&
      \bitbox{8}{\small 0}&
      \bitbox[]{8}{}\\
      \bitbox{8}{\small \(\Delta\)Time}&
      \bitbox{8}{\small Note}&
      \bitbox{8}{\small Velocity}&
      \bitbox[]{8}{}\\
      \bitbox{8}{\small \(\Delta\)Time}&
      \bitbox{8}{\small Note}&
      \bitbox{8}{\small 0}&
      \bitbox[]{8}{}\\
      \bitbox[]{32}{\(\vdots\)}
      \end{bytefield} 
\end{minipage}
    \caption{Running Status}
    \label{fig:running-status}
\end{figure}

\section{Simplifying the MIDI files}

To not have to decode every single MIDI message, most of which are not valuable to the \gls{midi}-Interrupter, a tool called \emph{tinymid}\sidenotes{Its source code can be found in QR code \newqrcode{https://github.com/marchersimon/tinymid}{tinymid}.} was used. It removes all unnecessary events, moves them to channel 0, converts them to running status if possible, and most importantly, it merges all tracks of a multiple track file into a single track to make it a single track file. For this, the whole file has to be decoded at once and all events sorted by their absolute delta time, which would have been much harder to do with the limited libraries and debugging capabilities available for AVR C++.